---
title: "Executive Summary"
format:
  html:
    page-layout: full
    toc: false
---

```{r setup}
#| include: false
library(dplyr); library(tidyr); library(readr); library(plotly)
library(lubridate); library(scales); library(stringr); library(knitr)

clean        <- read_csv("data/cleaned.csv.gz", show_col_types = FALSE)
yearly_nat   <- read_csv("data/yearly_national.csv", show_col_types = FALSE)
yearly_state <- read_csv("data/yearly_state.csv", show_col_types = FALSE)
state_sum    <- read_csv("data/state_summary.csv", show_col_types = FALSE)
city_sum     <- read_csv("data/city_summary.csv", show_col_types = FALSE)
monthly_nat  <- read_csv("data/monthly_national.csv", show_col_types = FALSE)

POLLUTANTS <- c("NO2","O3","SO2","CO")
COLOR <- c(NO2="#1F4E79", O3="#ED7D31", SO2="#7030A0", CO="#C00000")

# Headline numbers
get_change <- function(col, df = yearly_nat) {
  early <- df[[col]][df$Year == 2000]
  late  <- df[[col]][df$Year == 2015]
  round((late - early) / early * 100, 1)
}
no2_pct <- get_change("NO2 AQI")
o3_pct  <- get_change("O3 AQI")
so2_pct <- get_change("SO2 AQI")
co_pct  <- get_change("CO AQI")
```

::: {.hero}
::: {.hero-eyebrow}
Air Quality Performance Dashboard · 2000–2016
:::
<div class="hero-title">Three out of <span class="accent">Four</span></div>
::: {.hero-subtitle}
Sixteen years of EPA monitoring data across 47 states and 204 sites tell a clear policy story — and reveal the one pollutant that refused to fall.
:::
:::

## Performance Indicators {.unnumbered}

::: {.eyebrow}
National AQI change · 2000 → 2015
:::

::: {.kpi-grid}

::: {.kpi-card .kpi-good}
::: {.kpi-label}NO₂ Annual Mean AQI:::
::: {.kpi-value}`r no2_pct`%:::
::: {.kpi-delta .down}↓ Vehicle exhaust regulations:::
:::

::: {.kpi-card .kpi-good}
::: {.kpi-label}SO₂ Annual Mean AQI:::
::: {.kpi-value}`r so2_pct`%:::
::: {.kpi-delta .down}↓ Coal-plant retirements:::
:::

::: {.kpi-card .kpi-good}
::: {.kpi-label}CO Annual Mean AQI:::
::: {.kpi-value}`r co_pct`%:::
::: {.kpi-delta .down}↓ Catalytic converters:::
:::

::: {.kpi-card .kpi-alert}
::: {.kpi-label}O₃ Annual Mean AQI:::
::: {.kpi-value}+`r o3_pct`%:::
::: {.kpi-delta .up}↑ The chemistry holdout:::
:::

:::

::: {.eyebrow}
Dataset coverage
:::

::: {.stat-row}
::: {.stat}
<div class="num">1.75M</div>
<div class="lbl">Raw observations</div>
:::
::: {.stat}
<div class="num">412K</div>
<div class="lbl">After dedup</div>
:::
::: {.stat}
<div class="num">204</div>
<div class="lbl">Monitoring sites</div>
:::
::: {.stat}
<div class="num">144</div>
<div class="lbl">U.S. cities</div>
:::
::: {.stat}
<div class="num">47</div>
<div class="lbl">States covered</div>
:::
::: {.stat}
<div class="num">17</div>
<div class="lbl">Years of data</div>
:::
:::

## Headline Trend {.unnumbered}

::: {.chart-card}
::: {.chart-title}National Annual Mean AQI by Pollutant:::
::: {.chart-subtitle}Three pollutants improved sharply; ozone moved sideways. Hover any line for year-by-year values.:::

```{r headline-trend}
#| fig-height: 4.5
plot_ly() |>
  add_lines(x = yearly_nat$Year, y = yearly_nat$`NO2 AQI`,
            name = "NO2", line = list(color = COLOR["NO2"], width = 3, shape = "spline"),
            marker = list(size = 6, color = COLOR["NO2"]), mode = "lines+markers",
            hovertemplate = "<b>NO2</b><br>%{x}: %{y:.1f} AQI<extra></extra>") |>
  add_lines(x = yearly_nat$Year, y = yearly_nat$`O3 AQI`,
            name = "O3", line = list(color = COLOR["O3"], width = 3, shape = "spline"),
            marker = list(size = 6, color = COLOR["O3"]), mode = "lines+markers",
            hovertemplate = "<b>O3</b><br>%{x}: %{y:.1f} AQI<extra></extra>") |>
  add_lines(x = yearly_nat$Year, y = yearly_nat$`SO2 AQI`,
            name = "SO2", line = list(color = COLOR["SO2"], width = 3, shape = "spline"),
            marker = list(size = 6, color = COLOR["SO2"]), mode = "lines+markers",
            hovertemplate = "<b>SO2</b><br>%{x}: %{y:.1f} AQI<extra></extra>") |>
  add_lines(x = yearly_nat$Year, y = yearly_nat$`CO AQI`,
            name = "CO", line = list(color = COLOR["CO"], width = 3, shape = "spline"),
            marker = list(size = 6, color = COLOR["CO"]), mode = "lines+markers",
            hovertemplate = "<b>CO</b><br>%{x}: %{y:.1f} AQI<extra></extra>") |>
  layout(
    paper_bgcolor = "white", plot_bgcolor = "white",
    xaxis = list(title = "", gridcolor = "#E5E9ED", tickfont = list(family = "Source Sans 3")),
    yaxis = list(title = "Mean AQI", gridcolor = "#E5E9ED",
                 titlefont = list(family = "Source Sans 3", size = 12, color = "#5B6677")),
    hovermode = "x unified",
    legend = list(orientation = "h", y = -0.15, x = 0.5, xanchor = "center"),
    margin = list(l = 50, r = 30, t = 10, b = 50)
  ) |> config(displayModeBar = FALSE)
```
:::

::: {.insight}
::: {.insight-label}Key Insight:::
Where pollution had a clear source and a regulatory tool that addressed it directly — scrubbers for SO₂, catalytic converters for CO, vehicle standards for NO₂ — pollution fell sharply. Ozone has no single source, and didn't fall.
:::

## Geographic Distribution {.unnumbered}

::: {.two-col}

::: {.chart-card}
::: {.chart-title}Top 10 States by Average Air Pollution:::
::: {.chart-subtitle}Mean daily peak-pollutant AQI, all years pooled:::

```{r top-states}
#| fig-height: 4.5
top10_states <- state_sum |> arrange(desc(Max_AQI)) |> head(10) |> arrange(Max_AQI)
plot_ly(top10_states, y = ~State, x = ~Max_AQI, type = "bar",
        orientation = "h",
        marker = list(color = "#1F4E79"),
        text = ~sprintf("%.1f", Max_AQI), textposition = "outside",
        hovertemplate = "<b>%{y}</b><br>Mean Max AQI: %{x:.1f}<extra></extra>") |>
  layout(
    paper_bgcolor = "white", plot_bgcolor = "white",
    xaxis = list(title = "Mean AQI", gridcolor = "#E5E9ED"),
    yaxis = list(title = "", automargin = TRUE),
    margin = list(l = 100, r = 50, t = 10, b = 40)
  ) |> config(displayModeBar = FALSE)
```
:::

::: {.chart-card}
::: {.chart-title}Top 10 Most Polluted Cities:::
::: {.chart-subtitle}Sites with at least one year of observations:::

```{r top-cities}
#| fig-height: 4.5
top10_cities <- city_sum |> arrange(desc(Max_AQI)) |> head(10) |> arrange(Max_AQI)
plot_ly(top10_cities, y = ~paste0(City, ", ", State), x = ~Max_AQI, type = "bar",
        orientation = "h",
        marker = list(color = "#ED7D31"),
        text = ~sprintf("%.1f", Max_AQI), textposition = "outside",
        hovertemplate = "<b>%{y}</b><br>Mean Max AQI: %{x:.1f}<extra></extra>") |>
  layout(
    paper_bgcolor = "white", plot_bgcolor = "white",
    xaxis = list(title = "Mean AQI", gridcolor = "#E5E9ED"),
    yaxis = list(title = "", automargin = TRUE),
    margin = list(l = 140, r = 50, t = 10, b = 40)
  ) |> config(displayModeBar = FALSE)
```
:::

:::

::: {.insight}
::: {.insight-label}Pattern:::
Southern California cities (Fontana, Rubidoux, Fresno, Victorville) dominate the city ranking — basin geography traps pollutants against the surrounding mountains. The Eastern coal-belt states (Pennsylvania, Indiana, Kentucky, Ohio) dominate the state ranking due to legacy SO₂ emissions.
:::

## Seasonal Patterns {.unnumbered}

::: {.chart-card}
::: {.chart-title}Average AQI by Month — Pooled Across All Years:::
::: {.chart-subtitle}NO₂ and CO peak in winter (cold engines, heating). O₃ peaks in summer (photochemistry).:::

```{r seasonal}
#| fig-height: 4.5
m <- monthly_nat |>
  mutate(Month = month(YearMonth)) |>
  group_by(Month) |>
  summarise(across(ends_with("AQI"), \(x) mean(x, na.rm = TRUE)), .groups = "drop")

month_names <- c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")

plot_ly() |>
  add_lines(x = month_names, y = m$`NO2 AQI`, name = "NO2",
            line = list(color = COLOR["NO2"], width = 3, shape = "spline"),
            marker = list(size = 7), mode = "lines+markers",
            hovertemplate = "<b>NO2</b><br>%{x}: %{y:.1f}<extra></extra>") |>
  add_lines(x = month_names, y = m$`O3 AQI`, name = "O3",
            line = list(color = COLOR["O3"], width = 3, shape = "spline"),
            marker = list(size = 7), mode = "lines+markers",
            hovertemplate = "<b>O3</b><br>%{x}: %{y:.1f}<extra></extra>") |>
  add_lines(x = month_names, y = m$`SO2 AQI`, name = "SO2",
            line = list(color = COLOR["SO2"], width = 3, shape = "spline"),
            marker = list(size = 7), mode = "lines+markers",
            hovertemplate = "<b>SO2</b><br>%{x}: %{y:.1f}<extra></extra>") |>
  add_lines(x = month_names, y = m$`CO AQI`, name = "CO",
            line = list(color = COLOR["CO"], width = 3, shape = "spline"),
            marker = list(size = 7), mode = "lines+markers",
            hovertemplate = "<b>CO</b><br>%{x}: %{y:.1f}<extra></extra>") |>
  layout(
    paper_bgcolor = "white", plot_bgcolor = "white",
    xaxis = list(title = "", gridcolor = "#E5E9ED"),
    yaxis = list(title = "Mean AQI", gridcolor = "#E5E9ED"),
    hovermode = "x unified",
    legend = list(orientation = "h", y = -0.18, x = 0.5, xanchor = "center"),
    margin = list(l = 50, r = 30, t = 10, b = 60)
  ) |> config(displayModeBar = FALSE)
```
:::

## Strategic Recommendations {.unnumbered}

::: {.two-col}

::: {.chart-card}
**1. Continue source-targeted regulation.**
The 83% drop in SO₂ and 55% drop in CO are direct results of source-targeted rules. The strong NO₂↔CO correlation (r ≈ 0.62) means tailpipe regulation is a two-for-one investment.
:::

::: {.chart-card}
**2. Address ozone with a different playbook.**
Indirect formation means cutting precursors alone hasn't worked. Strategies need to combine continued NOₓ/VOC reduction with climate adaptation, particularly for hot summer days.
:::

::: {.chart-card}
**3. Geographic targeting.**
Risk is concentrated in basins (LA, Central Valley) and the legacy coal belt. National averages hide this — local interventions need local data.
:::

::: {.chart-card}
**4. Season-targeted public-health alerts.**
Winter: emphasize NO₂/CO (traffic + heating). Summer: emphasize O₃, especially on hot days for cardiopulmonary-sensitive groups.
:::

:::

---

::: {.eyebrow}
Continue Reading
:::

- **[Detailed Analysis →](analysis.qmd)** — Full statistical breakdown, distributions, correlations, formal tests
- **[Interactive Dashboard →](dashboard.qmd)** — Filter by state, year, pollutant, season — explore the data yourself
- **[Methodology →](methodology.qmd)** — How we cleaned, aggregated, and validated 1.75 million rows
